Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-01-13 paper-1/manuscript.tex
Editing

Based on these measures, we then asked why there were more spikes in \dev{} trials. We first focused on the contribution of $\T{}{}$ alone in an effort to confirm that thresholds are lower in \dev{}, and that this reduction in thresholds causes increased firing. We correlated the contrast in $\T{}{}$ at the start of target trials, i.e., $\Delta \T{A}{i} = \T[dev]{A}{i} - \T[msc]{A}{i}$ corresponding to the leftmost datapoints of the contrast histogram in Figure \ref{fig4}b, with the target trial response size contrast $\Delta \R{A}{i}$, across excitatory neurons. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in \dev{} than \msc{} (T = 2.69e+04, p = 1.58e-92, median = -0.166 mV; one-sided Wilcoxon, n = 800 excitatory neurons), and that this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 1.6e-36; one-sided Wald test, n = 800). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: The median of $\T{A}{i}$ among excitatory neurons was lower in \dev{} than \msc{} (T = 514, p = 1.67e-16, grand median = -0.094 mV; one-sided Wilcoxon, n = 120), indicating that networks were less adapted during oddball sequences. In addition, the correlation across excitatory neurons between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$ was almost universally negative (T = 40, p = 2.68e-21, median = -0.357). This indicates that neurons with a greater decrease in $\T[dev]{A}{}$ tended to respond more strongly to target stimuli.
% TODO: edit final sentence to end with ... tended to show a greater increase in their response to target stimuli.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig5.pdf}
    \caption{}
    \label{fig5}
\end{figure*}
\textbf{Fig. 5} Lower $\T[dev]{}{}$ is a consequence of lower \dev{} non-target response magnitude.
\textbf{a} $\T{}{}$ at the start of A trials (mean over trials) in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages in the \dev{} sequence ($\T[dev]{A}{i}$) and \msc{} sequence ($\T[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \T{A}{i}$). A and B stimulus locations for this and following figures are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the \msc{} sequence.
% TODO: Drop "for this .. figures" since they're everywhere now.
\textbf{b} Average response (in spikes per trial) to non-target trials ($\R{N}{i}$, with $N$ meaning B in \dev{}, and B, C, D, and E in \msc{}). Columns are arranged as in panel a.
\textbf{c} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$, and associated linear regression.
\textbf{d} $\Delta \R{N}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \R{N}{i}$ vs $\Delta \T{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations

>>>>

Based on these measures, we then asked why there were more spikes in \dev{} trials. We first focused on the contribution of $\T{}{}$ alone in an effort to confirm that thresholds are lower in \dev{}, and that this reduction in thresholds causes increased firing. We correlated the contrast in $\T{}{}$ at the start of target trials, i.e., $\Delta \T{A}{i} = \T[dev]{A}{i} - \T[msc]{A}{i}$ corresponding to the leftmost datapoints of the contrast histogram in Figure \ref{fig4}b, with the target trial response size contrast $\Delta \R{A}{i}$, across excitatory neurons. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in \dev{} than \msc{} (T = 2.69e+04, p = 1.58e-92, median = -0.166 mV; one-sided Wilcoxon, n = 800 excitatory neurons), and that this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 1.6e-36; one-sided Wald test, n = 800). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: The median of $\T{A}{i}$ among excitatory neurons was lower in \dev{} than \msc{} (T = 514, p = 1.67e-16, grand median = -0.094 mV; one-sided Wilcoxon, n = 120), indicating that networks were less adapted during oddball sequences. In addition, the correlation across excitatory neurons between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$ was almost universally negative (T = 40, p = 2.68e-21, median = -0.357). This indicates that neurons with a greater decrease in $\T[dev]{A}{i}$ tended to show a greater increase in their response to target stimuli.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig5.pdf}
    \caption{}
    \label{fig5}
\end{figure*}
\textbf{Fig. 5} Lower $\T[dev]{}{}$ is a consequence of lower \dev{} non-target response magnitude.
\textbf{a} $\T{}{}$ at the start of A trials (mean over trials) in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages in the \dev{} sequence ($\T[dev]{A}{i}$) and \msc{} sequence ($\T[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \T{A}{i}$). A and B stimulus locations are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the \msc{} sequence.
\textbf{b} Average response (in spikes per trial) to non-target trials ($\R{N}{i}$, with $N$ meaning B in \dev{}, and B, C, D, and E in \msc{}). Columns are arranged as in panel a.
\textbf{c} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$, and associated linear regression.
\textbf{d} $\Delta \R{N}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \R{N}{i}$ vs $\Delta \T{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations
